\section{Results}
\subsection{The mGPS Tool Implementation}
\begin{enumerate}
    \item \textbf{Goal of mGPS:} The goal is to predict the geographic origin of microbiomr samples using microbial relateve sequence abundances (RSA) and metadata.
    \item \textbf{Key Features of mGPS:} It does not require users to tweak complex setting. It mainly focuses on Geographically Informative Taxa.
    These are sepcific microbial taxa that show unique patterns in certain regions, making them useful for location predictions. 
    \item \textbf{Process for building the model:}
    \begin{enumerate}
        \item \textbf{Selecting the Best Taxa:}
        \begin{enumerate}
            \item A method called \textbf{recursive feautre elimination} is used to rank microbial taxa based on how useful they are for predicting geographic locations.
            \item Only the most informative taxa (GITs) are kept for building the model. 
        \end{enumerate}
        \item \textbf{Training the model:}
        \begin{enumerate}
            \item The model uses RSA data (the proportion of sequences belonging to each taxon).
            \item It starts by predicting broad regions like continents.
            \item Each prediction is then added as extra input to train models for finer levels (e.g., countries, latititude and longitutde) 
            \item This step-by-step approach helps the model to get better at predicting exact coordinates. 
        \end{enumerate}
        \item \textbf{Final Output:} The model predicts the latititude and longitutde of the microbiome samples.
    \end{enumerate}
    \item \textbf{Testing mGPS:} To check how well mGPS works, it was tested on different types of microbiome data.
    \begin{enumerate}
        \item Urban biome: Data from citites around the world. (Next-generation data)
        \item Soil biome: Data from soil microbes collected globally. (16srRNA data)
        \item Marine biome: Data from ocean microbes. (Shotgun Sequencing data)
    \end{enumerate}
    \item \textbf{Ensuring Reliability:} Data came from different sequencing methods and platforms, esuring the tool is not tied to specific ways of collecting of proecessing data.
    The rigorous testing showed that mGPS works across various datasets.
\end{enumerate}